---
title: "Data & Downloads"
subtitle: "Tidy, long-format versions of every figure on this site, with a full data dictionary."
toc-depth: 3
---

```{r}
#| label: setup
#| include: false
source(here::here("_common.R"))
meta <- read_meta()
```

EPA publishes this indicator's data as three spreadsheet-shaped CSV files, each
wrapped in a five-line metadata preamble and encoded in Windows-1252. They are
built for opening in Excel, not for analysis.

The files below carry the same numbers in long format, one observation per row,
UTF-8, with a machine-readable key for every series. Values are copied across as
text, so the source file's precision survives exactly. The original EPA files are
kept alongside them, unmodified, so any transformation can be checked.

```{r}
#| label: downloads
#| echo: false
rows <- lapply(meta$datasets, function(d) {
  data.frame(
    Figure   = d$figure,
    Tidy     = sprintf("[`%s`](data/%s)", d$file, d$file),
    Rows     = d$rows,
    Original = sprintf("[`%s`](data-raw/%s)", d$source_file, d$source_file),
    Updated  = d$web_update,
    check.names = FALSE
  )
})
knitr::kable(do.call(rbind, rows), row.names = FALSE, align = "llrll")
```

Also in the repository: [`data/meta.yml`](data/meta.yml), the machine-readable
version of the dictionary below, and
[`data-raw/PROVENANCE.md`](data-raw/PROVENANCE.md), which records where each
source file came from and its checksum.

## Two things worth knowing before you use these

### Suppressed values are not zeros

CDC withholds a rate when the underlying death count falls below its disclosure
threshold, writing `Suppressed` in place of a number. This happens twice in
Figure 2, both in the non-Hispanic Black series.

In these files such an observation keeps its row, carries an empty `value`, and
sets `flag` to `suppressed`. It means "too few deaths to publish", which is
different from zero and different from missing. Charts on this site draw a gap
there. EPA's own published Figure 2 plots those two points at zero, which reads
as though nobody died.

### The 1998/1999 break is real

Causes of death were classified under ICD-9 through 1998 and ICD-10 from 1999.
Rates either side of that change are not directly comparable, which is why EPA
draws Figure 1's underlying-cause line with a gap.

EPA records this only as a visual break and a footnote. Here it is a column,
`icd_revision`, so the discontinuity is visible to anything reading the file and
a chart cannot accidentally interpolate across it.

## Data dictionary

```{r}
#| label: dictionary
#| echo: false
#| output: asis
for (d in meta$datasets) {
  cat("\n### ", d$figure, ": `", d$file, "`\n\n", sep = "")
  cat("> ", d$figure_title, "\n\n", sep = "")

  facts <- data.frame(
    Field = c("Rows", "Unit", "Data source", "Web update", "Built from", "Source encoding"),
    Value = c(as.character(d$rows), d$unit, d$data_source, d$web_update,
              paste0("`", d$source_file, "`"), d$source_encoding),
    check.names = FALSE
  )
  print(knitr::kable(facts, row.names = FALSE))
  cat("\n\n")

  cols <- do.call(rbind, lapply(d$columns, function(cc) {
    data.frame(Column = paste0("`", cc$name, "`"), Type = cc$type,
               Description = cc$description, check.names = FALSE)
  }))
  print(knitr::kable(cols, row.names = FALSE))
  cat("\n\n")

  if (!is.null(d$series) && length(d$series)) {
    cat("**Series**\n\n")
    ser <- do.call(rbind, lapply(d$series, function(s) {
      data.frame(
        Key      = paste0("`", s$key, "`"),
        Label    = s$label,
        Coverage = if (is.null(s$coverage)) "" else s$coverage,
        `Source column` = paste0("`", s$source_header, "`"),
        check.names = FALSE
      )
    }))
    print(knitr::kable(ser, row.names = FALSE))
    cat("\n\n")
    for (s in d$series) {
      if (!is.null(s$note)) cat("- `", s$key, "`: ", s$note, "\n", sep = "")
    }
    cat("\n")
  }
  if (!is.null(d$note)) cat("\n", d$note, "\n\n", sep = "")
}
```

## Rights

```{r}
#| label: rights
#| echo: false
#| output: asis
cat(meta$indicator$rights, "\n\n")
cat("Original indicator page: <", meta$indicator$source_page, ">\n\n", sep = "")
cat("Technical documentation: <", meta$indicator$technical_documentation, ">\n", sep = "")
```
