---
title: "Heat-Related Deaths"
subtitle: "This indicator presents data on deaths classified as \"heat-related\" in the United States."
---

```{r}
#| label: setup
#| include: false
source(here::here("_common.R"))
source(here::here("R/figures.R"))
meta <- read_meta()
```

```{r}
#| label: caption-block
#| include: false
# Renders a figure's caption strip from data/meta.yml: title, data source, and
# web-update date, all copied verbatim from the CSV EPA published. Nothing here
# is hand-typed, so it cannot drift from the source file.
caption_block <- function(file) {
  m <- meta_for(file, meta)
  knitr::asis_output(sprintf(
    "**%s**\n\nData source: %s <br> Web update: %s\n",
    m$figure_title, m$data_source, m$web_update
  ))
}
```

## Figure 1

```{r}
#| label: fig-1-caption
#| echo: false
#| output: asis
caption_block("heat_deaths_annual.csv")
```

```{r}
#| label: fig-1
#| echo: false
fig_1()
```

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-1-table
#| echo: false
knitr::kable(fig_1_table(), digits = 3)
```

Download: [tidy CSV](data/heat_deaths_annual.csv) · [original EPA file](data-raw/heat-deaths_fig-1.csv)

</details>

## Figure 2

```{r}
#| label: fig-2-caption
#| echo: false
#| output: asis
caption_block("heat_deaths_summer_cvd.csv")
```

```{r}
#| label: fig-2
#| echo: false
fig_2()
```

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-2-table
#| echo: false
knitr::kable(fig_2_table(), digits = 3)
```

Download: [tidy CSV](data/heat_deaths_summer_cvd.csv) · [original EPA file](data-raw/heat-deaths_fig-2.csv)

</details>

## Example: The 1995 Chicago Heat Wave

```{r}
#| label: fig-example-caption
#| echo: false
#| output: asis
caption_block("chicago_1995_heat_wave.csv")
```

```{r}
#| label: fig-example
#| echo: false
fig_example()
```

<details><summary>Show the data behind this figure</summary>

```{r}
#| label: fig-example-table
#| echo: false
knitr::kable(utils::head(fig_example_table(), 10), digits = 1,
            caption = "First 10 of 92 days.")
```

Download: [tidy CSV](data/chicago_1995_heat_wave.csv) · [original EPA file](data-raw/heat-deaths_example.csv)

</details>

::: {.callout-note appearance="simple" icon=false}
## 🚧 Coming soon

The rest of this page (Key Points, Background, About the Indicator, Indicator
Notes, Data Sources, and References) is EPA's own published text and is being
added next, extracted directly from EPA's source document rather than retyped.
See [Data & Downloads](data.qmd) for what's already here: tidy data, a full
dictionary, and both departures from EPA's published charts.
:::
